\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-14 chapters/gaze_independence.tex
Self-plagiarism check
Started processing comments MDV

@@

\subsection{Gaze impairment and \acl{lis}}

\newcommand\fnlis{\footnote{Multiple definitions of \ac{lis} are encountered in
\ac{bci} and neurological literature.
Some definitions include only those with tetraplegy without eye movements used
for communications.
Others distinguish Complete Locked-in Syndrome (CLIS) with full body paralysis,
including no eye motor control at all, from a \ac{lis} state with some preserved eye
movements or minor motor output.
While some definitions only include stroke or \ac{tbi} with damage to
specific regions in the brain (midbrain, brainstem, or
cerebellum)~\cite{Smith2005}, it can also generally refer to the state of full body paralysis
or loss of muscle tone incurred in neurodegenerative diseases, combined with
the inability to speak, such as occurs in late-stage ALS.}}

@@

We believe it is generally not opportune to delve too deeply into the underlying
ophthalmological and neurological mechanisms for each of the specific
conditions, or even the exact symptoms.
In clinical reality, every \ac{bci} user has different symptoms, which lead to their own
unique set of preserved capabilities and visual skills.
From a solution-oriented \ac{bci} engineering point of view, the etiology of the symptoms can be abstracted away.
Instead of categorizing users by their etiology, we will refer to those who might benefit from
\ac{bci} assistive communication technology as having \ac{sspi}, in line with
the terminology used by~\textcite{FriedOken2020}.
If they have eye motor impairment that affects their use of visual \ac{bci} or
eye-tracking solutions, we use the term \ac{sspgi} (see \cref{fig:gaze_independence/venn}).

\begin{figure}
  \input{figures/gaze_independence/lis_venn.tikz.tex}
  \caption[Venn diagram of potential users of \acs{bci} assistive technology.]{%
    Venn diagram of potential users of assistive technology.
    Patients with \acf{sspi} are the general target population.
    If eye gaze-based solutions are not properly suited for an individual,
    they have \acf{sspgi}.
    Individuals with \ac{lis} have no muscular control left, or it is very limited,
    such as the ability to communicate through a binary switching system.
    \Acp{bci} are especially relevant for the last two groups, as they are not
    able to use eye tracking solutions.
  }
  \label{fig:gaze_independence/venn}
\end{figure}

@@

\section{Research objectives}

>>>>

\noindent\emph{The last paragraph of \cref{sec:gaze-independence/sota},
\cref{sec:gaze-independence/oculomotor/gaze}, and the first paragraph of
\cref{sec:gaze-independence/objectives} were adapted from
\textcite{VanDenKerchove2024}.}

@@

\subsection{Gaze impairment and \acl{lis}}
\label{sec:gaze-independence/oculomotor/gaze}
\newcommand\fnlis{\footnote{Multiple definitions of \ac{lis} are encountered in
\ac{bci} and neurological literature.
Some definitions include only those with tetraplegy without eye movements used
for communications.
Others distinguish Complete Locked-in Syndrome (CLIS) with full body paralysis,
including no eye motor control at all, from a \ac{lis} state with some preserved eye
movements or minor motor output.
While some definitions only include stroke or \ac{tbi} with damage to
specific regions in the brain (midbrain, brainstem, or
cerebellum)~\cite{Smith2005}, it can also generally refer to the state of full body paralysis
or loss of muscle tone incurred in neurodegenerative diseases, combined with
the inability to speak, such as occurs in late-stage ALS.}}

@@

We believe it is generally not opportune to delve too deeply into the underlying
ophthalmological and neurological mechanisms for each of the specific
conditions, or even the exact symptoms.
In clinical reality, every \ac{bci} user has different symptoms, which lead to their own
unique set of preserved capabilities and visual skills.
From a solution-oriented \ac{bci} engineering point of view, the etiology of the symptoms can be abstracted away.
Instead of categorizing users by their etiology, we will refer to those who might benefit from
\ac{bci} assistive communication technology as having \ac{sspi}, in line with
the terminology used by~\textcite{FriedOken2020}.
If they have eye motor impairment that affects their use of visual \ac{bci} or
eye-tracking solutions, we use the term \ac{sspgi} (see \cref{fig:gaze-independence/venn}).

\begin{figure}
  \input{figures/gaze_independence/lis_venn.tikz.tex}
  \caption[Venn diagram of potential users of \acs{bci} assistive technology.]{%
    Venn diagram of potential users of assistive technology.
    Patients with \acf{sspi} are the general target population.
    If eye gaze-based solutions are not properly suited for an individual,
    they have \acf{sspgi}.
    Individuals with \ac{lis} have no muscular control left, or it is very limited,
    such as the ability to communicate through a binary switching system.
    \Acp{bci} are especially relevant for the last two groups, as they are not
    able to use eye tracking solutions.
  }
  \label{fig:gaze-independence/venn}
\end{figure}

@@

\section{Research objectives}
\label{sec:gaze-independence/objectives}

### 2024-11-14 chapters/patients.tex
Self-plagiarism check
Started processing comments MDV

\begin{table}[t]
  \centering
  \footnotesize
  \input{chapters/patients_table_eye.tex}
  \caption[Visual skills of the included participants.]{%
  Visual skills of the included participants.
  Visual~\ac{bci} skills~\cite{FriedOken2020}
  were assessed with a combination of self-reported issues
  by the subject and the NeuroEye~\cite{Hassan2022} test. \BigCross\
  impaired, \BigDiamondshape\ severely impaired.
  logMAR: lower is better.}
  \label{tab:patients/eye}
\end{table}
\todo{where these values corrected for distance?}

@@

\subsection{\Ac{bci} stimulation}

@@

\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/patients/fig_gaze.png}
}{%
  Distribution of the recorded gaze position.
}{%
  Distribution of the recorded gaze position during the experimental session in the three \ac{vsa}
  conditions.
  Crosshairs represent stimulus positions, with the orange ones cued during
  the given condition.
  Subjects PB2 and PC4 preferred covert \ac{bci} operation, with PB2 resting gaze
  near the middle of the screen, and PC4 near the bottom.
}{fig:patients/gaze}

@@

\begin{figure}[t]
  \input{figures/patients/fig_decode.pgf}
  \caption[%
    Decoding performance in different \acs{vsa} settings.
  ]{%
    Decoding performance in different \ac{vsa} settings reported as
    single-trial \ac{rocauc}.
    Free \ac{vsa} is generally on par with performance in the overt \ac{vsa}
    setting.
    Performance in the covert \ac{vsa} setting with central gaze fixation is lower, but can
    be improved with the \ac{wcble} decoder.
    95\% confidence intervals were calculated using 1000 bootstrapping
    repetitions.
  }
  \label{fig:patients/decode}
\end{figure}

@@

@@

\begin{figure}[t]
  \input{figures/patients/fig_cross.pgf}
  \caption[Cross-setting calibration and decoding performance.]{
    Decoding performance when calibrating the \ac{tlda} decoder in a given \ac{vsa}
    setting, and evaluating it in another, reported as
    single-trial \ac{rocauc}.
    For participant PA1, free \ac{vsa} performance improved when
    calibrated with overt gaze fixation.
    For participants PA1, PB2, and PC3, performance in the covert \ac{vsa} setting with central gaze fixation
    improved when calibrating with overt gaze fixation.
  }
  \label{fig:patients/cross}
\end{figure}

>>>>

\begin{table}[t]
  \centering
  \footnotesize
  \input{chapters/patients_table_eye.tex}
  \caption[Visual skills of the included participants.]{%
  Visual skills of the included participants.
  Visual~\ac{bci} skills~\cite{FriedOken2020}
  were assessed with a combination of self-reported issues
  by the subject and the NeuroEye~\cite{Hassan2022} test. \BigCross\
  impaired, \BigDiamondshape\ severely impaired.
  logMAR: lower is better.}
  \label{tab:patients/eye}
\end{table}

@@

\subsection{\Acs{bci} stimulation}

@@

\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/patients/fig_gaze.png}
}{%
  \caption[Distribution of the recorded gaze position.]{%
    Distribution of the recorded gaze position during the experimental session in the three \ac{vsa}
    conditions.
    Crosshairs represent stimulus positions, with the orange ones cued during
    the given condition.
    Subjects PB2 and PC4 preferred covert \ac{bci} operation, with PB2 resting gaze
    near the middle of the screen, and PC4 near the bottom.
  }
  \label{fig:patients/gaze}
}

@@

@@

\begin{figure}
  \input{figures/patients/fig_decode.pgf}
  \caption[%
    Decoding performance in different \acs{vsa} settings.
  ]{%
    Decoding performance in different \ac{vsa} settings reported as
    single-trial \ac{rocauc}.
    Free \ac{vsa} is generally on par with performance in the overt \ac{vsa}
    setting.
    Performance in the covert \ac{vsa} setting with central gaze fixation is lower, but can
    be improved with the \ac{wcble} decoder.
    95\% confidence intervals were calculated using 1000 bootstrapping
    repetitions.
  }
  \label{fig:patients/decode}
\end{figure}

@@

\begin{figure}
  \input{figures/patients/fig_cross.pgf}
  \caption[Cross-setting calibration and decoding performance.]{
    Decoding performance when calibrating the \ac{tlda} decoder in a given \ac{vsa}
    setting, and evaluating it in another, reported as
    single-trial \ac{rocauc}.
    For participant PA1, free \ac{vsa} performance improved when
    calibrated with overt gaze fixation.
    For participants PA1, PB2, and PC3, performance in the covert \ac{vsa} setting with central gaze fixation
    improved when calibrating with overt gaze fixation.
  }
  \label{fig:patients/cross}
\end{figure}
